---
title: 'USDA Nutritional Data Analysis'
output:
  word_document: default
  pdf_document: default
---
<h2> Created by Christopher Wong: 8-4-22 </h2> 
<h3> Last updated: 1-19-25 </h3> 
---

#### Data Preprocessing
Read and merged the datasets on a common key, handled missing values, 
converted non-numeric data into numeric formats, created a cleaned dataset for analysis.
```{r}
# Read the data files into data frames
macroNut <- read.csv("./USDA_Macronutrients.csv", header = TRUE)
microNut <- read.csv("./USDA_Micronutrients.csv", header = TRUE)

# Merge the data frames using the variable "ID"
USDA <- merge(macroNut, microNut, by = "ID")

# Checking the data types
str(USDA)

# Delete commas from the sodium recs, then make data types numeric instead
USDA$Sodium <- gsub(",", "", USDA$Sodium)
USDA$Sodium <- as.numeric(USDA$Sodium)

# Delete commas from the potassium recs, then make data types numeric instead
USDA$Potassium <- gsub(",", "", USDA$Potassium)
USDA$Potassium <- as.numeric(USDA$Potassium)

# Remove records (rows) with missing values in more than 6 attributes (columns)
# get the rows with 4 or less missing values
USDA <- USDA[rowSums(is.na(USDA)) <= 4, ]

# Checking the data types to see # of records remaining in the data frame
str(USDA)

# Replace missing values for Sugar, Vitamin E and Vitamin D, with the mean value for the respective variable
USDA$Sugar[is.na(USDA$Sugar)]       <- mean(USDA$Sugar[!is.na(USDA$Sugar)])
USDA$VitaminE[is.na(USDA$VitaminE)] <- mean(USDA$VitaminE[!is.na(USDA$VitaminE)])
USDA$VitaminD[is.na(USDA$VitaminD)] <- mean(USDA$VitaminD[!is.na(USDA$VitaminD)])

#remove all remaining records with missing values, save to new data frame
USDAclean <- USDA[complete.cases(USDA), ]

# Checking the data types to see # of records remaining in the data frame
str(USDAclean)
```

#### Sodium Content Analysis
Consuming foods with such high sodium levels can lead to health issues such as hypertension, heart disease, and stroke. 
It is important to monitor and limit the intake of these foods, especially for individuals with health conditions that require low sodium diets.
Below is a list of the top 20 foods and ingredients with the highest sodium levels.
```{r}
# Sort the data frame by Sodium in descending order and select the top 20 rows
top20_sodium <- USDAclean[order(-USDAclean$Sodium), ][1:20, ]

# Display the top 20 foods/ingredients with the highest sodium levels
top20_sodium

```

#### Vitamin C distribution in foods. 
Vitamin C is an essential nutrient that plays a key role in maintaining overall health and immunity.
Below are histograms showing the distribution of Vitamin C content in foods. 
We see that the vast majority of foods have low to moderate levels of Vitamin C.
For a more detailed view, we also created a histogram for foods with Vitamin C levels of 100 or more.

```{r}
# Histogram for Vitamin C distribution for all foods
hist(USDAclean$VitaminC,
     breaks = seq(0, max(USDAclean$VitaminC, na.rm = TRUE), by = 100),
     xlab = "Vitamin C distribution in foods",
     main = "Histogram of Vitamin C Distribution",
     xlim = c(0, max(USDAclean$VitaminC, na.rm = TRUE)),
     xaxt = "n")  # Suppress the default x-axis
axis(1, at = seq(0, max(USDAclean$VitaminC, na.rm = TRUE), by = 100))

# Filter the data to include only foods with Vitamin C levels of 100 or more
high_vitaminC <- USDAclean[USDAclean$VitaminC >= 100, ]
# Histogram for Vitamin C distribution for foods with Vitamin C levels of 100 or more
hist(high_vitaminC$VitaminC,
     breaks = seq(100, max(high_vitaminC$VitaminC, na.rm = TRUE), by = 100),
     xlab = "Vitamin C distribution in foods (100 or more)",
     main = "Histogram of Vitamin C Distribution (100 or more)",
     xlim = c(100, max(high_vitaminC$VitaminC, na.rm = TRUE)),
     xaxt = "n")  # Suppress the default x-axis

# Customize the x-axis to reflect breaks of 100
axis(1, at = seq(100, max(high_vitaminC$VitaminC, na.rm = TRUE), by = 100))
```

#### Distribution of values for TotalFat, Protein and Carbohydrate.
The boxplot below visualizes the distribution of Total Fat, Protein, and Carbohydrates in the dataset. 
This visualization aids in comparing the central tendencies and variability of these macronutrients, identifying outliers, 
and gaining insights into the nutritional composition of the foods in the dataset.
```{r}
boxplot(USDAclean$TotalFat, USDAclean$Protein, USDAclean$Carbohydrate,
        main  = "Distribution of: Total Fat, Protein, and Carbohydrates.",
        names = c("Total Fat", "Protein", "Carbohydrates"))
```

#### Relationship between a food's TotalFat content and its Calorie content.
The graph below indicates that there is a positive relationship between Total Fat content and Calorie content in foods. 
This means that foods with higher fat content tend to have higher calorie content. 
This relationship is expected because fat is a dense source of energy, providing 9 calories per gram, 
compared to 4 calories per gram for protein and carbohydrates. 
Therefore, as the fat content in foods increases, the overall calorie content also increases.

Understanding this relationship is important for nutritional analysis and dietary planning, 
as it helps in assessing the energy contribution of fat in different foods.
```{r}
plot(USDAclean$TotalFat, USDAclean$Calories,
     main = "Relationship between Total Fat Content and Calorie Content.",
     xlab = "Total fat Content",
     ylab = "Calorie Content ")

```

#### Adding high nutrient indicators and counting foods with high sodium and fat
The analysis reveals that there are 1,827 foods in the dataset that have both high sodium and high fat content. 
This indicates a significant number of foods that may pose health risks due to their high levels of these two nutrients.

By adding binary indicators, we provide a clear way to identify and categorize foods based on their nutrient levels. 
This can be useful for further nutritional analysis and dietary planning.
```{r}
# Make new columns for: High Sodium, High Calories, High Protein, High Sugar, and High Fat
# Check if value is greater than the avg. if true: set it equal to 1
USDAclean$HSodium   <- 0
USDAclean$HSodium[USDAclean$Sodium     > mean(USDAclean$Sodium)]   <- 1

USDAclean$HCalories <- 0
USDAclean$HCalories[USDAclean$Calories > mean(USDAclean$Calories)] <- 1

USDAclean$HProtein  <- 0
USDAclean$HProtein[USDAclean$Protein   > mean(USDAclean$Protein)]  <- 1

USDAclean$HSugar    <- 0
USDAclean$HSugar[USDAclean$Sodium      > mean(USDAclean$Sodium)]   <- 1

USDAclean$HFat      <- 0
USDAclean$HFat[USDAclean$Sodium        > mean(USDAclean$Sodium)]   <- 1

# Number of foods have both high sodium and high fat
sum(apply(USDAclean[c("HSodium", "HFat")], 1, function(x) all(x == 1)))
# 1827 foods have both high sodium and high fat

```

#### Average amount of iron for high and low protein foods.
Iron is an essential nutrient that plays a crucial role in oxygen transport and energy production. 
Understanding the relationship between protein and iron content in foods can help in creating balanced diets that support overall health and well-being.

Findings:
- Low Protein Foods: The average amount of iron is approximately 2.70 mg.
- High Protein Foods: The average amount of iron is approximately 3.07 mg.

High protein foods tend to have a slightly higher average iron content compared to low protein foods. 
This information is important for nutritional analysis and dietary planning, 
helping to ensure adequate intake of both protein and iron.


```{r}
# average amount of iron, for high and low protein foods
meanIron.lowProtein  <- mean(USDAclean$Iron[USDAclean$HProtein == 0])
meanIron.lowProtein
meanIron.highProtein <- mean(USDAclean$Iron[USDAclean$HProtein == 1])
meanIron.highProtein
```

#### Custom HealthCheck program to detect unhealthy foods.
This section implements a custom HealthCheck program to classify foods as either "Pass" or "Fail" 
based on their sodium, sugar, and fat content. 
The results show that 1,827 foods fail the HealthCheck, 
highlighting the importance of monitoring and improving the nutritional quality of foods to support better health outcomes.
```{r}
# Used this algorithm chart as a basis for the function
require(jpeg)
file <- readJPEG("HealthCheck.jpg")
plot(1:4, ty = "n", ann = FALSE, xaxt = "n", yaxt = "n")
rasterImage(file, 1, 1, 4, 4)

#logic according to flowchart basis
HealthCheck <- function(sodium, sugar, fat) {

  if      (sodium == 0) return("Pass")

  else if (sugar == 0)  return("Pass")
  else if (fat == 0)    return("Pass")

  else                  return("Fail")
}
#input each type of food into the function
USDAclean$HealthCheck <- apply(USDAclean[, c("HSodium", "HSugar", "HFat")], 1,
                function(x) HealthCheck(x["HSodium"], x["HSugar"], x["HFat"]))

#number of failed foods
length(which(USDAclean$HealthCheck == "Fail"))


```

#### Correlation among Calories, Protein, Total Fat, Carbohydrate, Sodium and Cholesterol.
The plot below provides a visual representation of the correlation between Calories, Protein, Total Fat, Carbohydrate, Sodium, and Cholesterol.

- Positive Correlations: Strong positive correlations are indicated by darker colors and larger circles. For example, Calories show a strong positive correlations with Total Fat.
- Negative Correlations: Negative correlations, are indicated by different colors and smaller circles.
- Strength of Relationships: The size and color intensity of the circles indicate the strength of the correlations, with larger and darker circles representing stronger correlations.
This visualization helps in understanding how these nutritional components are related to each other, which is useful for nutritional analysis and dietary planning.
```{r}
library(corrplot)

#create matrix
nutCorrplot = cor(USDAclean[, c("Calories", "Protein", "TotalFat",
                  "Carbohydrate", "Sodium", "Cholesterol")])

#plotting
corrplot(nutCorrplot)

```

#### Statisical significance of the correlation between Calories and Total Fat.
The correlation between Calories and Total Fat is statistically significant (p-value < 2.2e-16), 
indicating a strong relationship between these two variables.
```{r}
# Find the pvalue using the function
cor.test(USDAclean$Calories, USDAclean$TotalFat)
# Set alpha = 0.05.
# Now we compare it to the pvalue <  2.2 E-16
# Since the pvalue is less than alpha, the correlation between Calories and Total Fat statistically significant.

```

#### Linear Regression Model for Calories based on Protein, Total Fat, Carbohydrate, Sodium, and Cholesterol.
This model is used to predict Calories based on the independent variables: Protein, Total Fat, Carbohydrate, Sodium, and Cholesterol.
This analysis helps in understanding the contribution of each macronutrient to the calorie content of foods.
Sodium has a p-value of 0.055, so it does not significantly contribute to the prediction of Calories.
```{r}
nutLm <- lm (Calories ~ Protein + TotalFat + Carbohydrate + Sodium + Cholesterol,
             data = USDAclean)
summary(nutLm)
# Sodium has a p-value of 0.05500, making it a non-significant variable.

```

#### New model using only significant independent variables.
```{r}
# Create a new model by using only the significant independent variables.
sigVarLm <- (lm(Calories ~ Protein + TotalFat + Carbohydrate + Cholesterol,
                data = USDAclean))
summary(sigVarLm)

```

#### Using the new model to predict number of Calories for a new product.
The new model predicts that a product with 0.1g Protein, 37g Total Fat, 400g Carbohydrate, and 75mg Cholesterol 
will have approximately 1827.4 Calories.
```{r}
# New product data
new_product <- data.frame(Protein = 0.1, TotalFat = 37, Carbohydrate = 400, Cholesterol = 75) 

# Predict the value for Calories using the new model
predicted_calories <- predict(sigVarLm, newdata = new_product)

# Display the predicted value
predicted_calories
# 1827.405 is the predicted value for Calories
```

#### If the new product's Carbohydrate count increases by 10000%, how much change will occur on Calories?
The Calories would increase by approximately 1083.33% when the Carbohydrate amount increases from 400 to 40000. 
This large increase is due to the direct proportional relationship between Carbohydrate and Calories as 
indicated by the linear regression model.
```{r}
# Extract the coefficient for Carbohydrate from the model
carb_coef <- coef(sigVarLm)["Carbohydrate"]

# Calculate the change in Carbohydrate
carb_change <- 40000 - 400

# Calculate the change in Calories
calories_change <- carb_coef * carb_change

# Calculate the initial predicted Calories for Carbohydrate = 400
initial_calories <- predict(sigVarLm, newdata = data.frame(Protein = 0.1, TotalFat = 37, Carbohydrate = 400, Cholesterol = 75))

# Calculate the new predicted Calories for Carbohydrate = 40000
new_calories <- predict(sigVarLm, newdata = data.frame(Protein = 0.1, TotalFat = 37, Carbohydrate = 40000, Cholesterol = 75))

# Calculate the percentage change in Calories
percent_change <- ((new_calories - initial_calories) / initial_calories) * 100

# Display the results
initial_calories
new_calories
percent_change
```
#### How do Protein, Total Fat, and Carbohydrate content relate to the Calorie content in foods?
The scatter plots and pair plot below will help us understand this relationship. 
The linear regression lines in the scatter plots will indicate the direction and strength of the relationships. 
The pair plot will provide a comprehensive view of the relationships between all four attributes.
```{r}
# Load necessary libraries
library(ggplot2)
library(GGally)

# Scatter plot for Protein vs Calories
ggplot(USDAclean, aes(x = Protein, y = Calories)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Protein vs Calories", x = "Protein", y = "Calories")

# Scatter plot for Total Fat vs Calories
ggplot(USDAclean, aes(x = TotalFat, y = Calories)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Total Fat vs Calories", x = "Total Fat", y = "Calories")

# Scatter plot for Carbohydrate vs Calories
ggplot(USDAclean, aes(x = Carbohydrate, y = Calories)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  labs(title = "Carbohydrate vs Calories", x = "Carbohydrate", y = "Calories")

# Pair plot to visualize relationships between Protein, Total Fat, Carbohydrate, and Calories
ggpairs(USDAclean, columns = c("Protein", "TotalFat", "Carbohydrate", "Calories"),
        title = "Pair Plot of Protein, Total Fat, Carbohydrate, and Calories")
```